D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-02 ootavacuous.tex
More Alan Stern feedback (December 2nd)
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	The result is that a correct C++ implementation necessarily
	preserves the semantic dependencies that prevent OOTA
	cycles from forming.
\end{abstract}

@@

We therefore define an OOTA cycle (also known as a causal cycle) as a
cycle in the memory-model sense involving the load-buffering (LB) pattern.
Within each thread in the cycle, the value stored depends on
the value loaded.
This dependency might be a data dependency
(see Listing~\ref{lst:OOTA Cycle})
or it might be a control dependency
(see Listing~\ref{lst:OOTA Cycle Based on Conditionals}).\footnote{
	Sufficiently motivated readers should have no problem constructing
	examples involving address dependencies.}
There must be at least two distinct values such that if a given load
returns a given value, that value forces the cycle to execute so as to
cause the corresponding store's value to be equal to that given value, and
for that store to occur early enough for the load to return that value.

@@

Because C++ volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2022N4910},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any C++ volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

Although non-volatile accesses to atomic objects are not observable behavior,
they are still subject to the memory model.
In particular, implementations must not behave as if the source code has
additional:

@@

@@

\paragraph{Omitted Atomic Stores}
In contrast, an implementation is permitted to omit atomic stores
under less restricted circumstances.
For example, a pair of back-to-back stores to \co{x} might always be
executed such that no other thread accesses \co{x} during the time
between those two stores.
This means that, short of inspecting the assembly code, the user has no
way of proving that the store was in fact omitted.
The user can always resort to volatile atomic stores or inline
assembly\footnote{
	For example, by placing the Linux-kernel \co{barrier()} macro
	between the two stores.
	This macro is an empty GCC \co{asm} that specifies the \co{memory}
	clobber.}
to prevent the compiler from omitting a store.
However, omitting a store cannot create an OOTA cycle.

@@

We are grateful to David Goldblatt and Jade Alglave for their careful
review of an early draft of this paper and to John Wikerson for asking
Paul for a rant and taking the proffered rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked an
insightful question at the right time, and for his ``Deathstation 9000''
demonic CPU.
Nonetheless, all errors and omissions in this paper are the sole property
of the authors.

>>>>

	The result is that a correct per-execution-path C++ implementation
	necessarily preserves the semantic dependencies that prevent
	OOTA cycles from forming.
	Implementations that undertake complex cross-thread analysis
	still need to be careful.
\end{abstract}

@@

We therefore define an OOTA cycle (also known as a causal cycle) as a
cycle in the memory-model sense involving only reads-from links, that
is, the value stored in on thread being read by the next thread.
Within each thread in the cycle, the value stored depends on
the value loaded.
This dependency might be a data dependency
(see Listing~\ref{lst:OOTA Cycle})
or it might be a control dependency
(see Listing~\ref{lst:OOTA Cycle Based on Conditionals}).\footnote{
	Sufficiently motivated readers should have no problem constructing
	examples involving address dependencies.}
@@@ There must be at least two distinct values such that if a given load
returns a given value, that value forces the cycle to execute so as to
cause the corresponding store's value to be equal to that given value, and
for that store to occur early enough for the load to return that value.

@@

Because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2022N4910},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

Although C++ non-volatile accesses to atomic objects are not observable
behavior, they are still subject to the memory model.
In particular, implementations must not behave as if the source code has
additional:

@@

The discussion of these two constraints is followed by a discussion of
when redundant atomic stores may safely be omitted.

@@

\paragraph{Omitted Redundant Atomic Stores}
In contrast, a pair of back-to-back non-volatile atomic stores to \co{x}
might always be executed such that no other thread accesses \co{x}
during the time between those two stores.
This means that, short of inspecting the assembly code, the user has no
way of proving that the C++ implementation acted as if that store was
omitted from the source code.
In cases where this is undesirable, the user can resort to volatile
atomic stores or to inline assembly\footnote{
	For example, by placing the Linux-kernel \co{barrier()} macro
	between the two stores.
	This macro is an empty GCC \co{asm} that specifies the \co{memory}
	clobber.}
to prevent the compiler from omitting a store.
However, acting as if such a redundant store was omitted from the source
code cannot create an OOTA cycle.
After all, the implementation is already within its as-if rights to
assume that the first of a back-to-back pair of non-volatile relaxed
atomic stores was omitted from the source code.

@@

We are grateful to David Goldblatt, Jade Alglave, and Alan Stern for their
careful review of an early draft of this paper and to John Wikerson for
asking Paul for a rant and taking the proffered rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked an
insightful question at the right time, and for his ``Deathstation 9000''
demonic CPU.
Nonetheless, all errors and omissions in this paper are the sole property
of the authors.
